Communicative pressures at the semantics-pragmatics interface:\\ Learning biases may prevent the lexicalization of pragmatic inferences

### 2016-09-21 paper/paper.tex
added cogsci files, extended draft section 3

We see two main explanations for the lack of upper-bounds in the lexical meaning of weak scalar expressions. The first is that their truth-conditional compatibility with stronger expressions lends them a broader range of application. When the speaker is unwilling or unable to assert a stronger alternative, e.g. because she lacks knowledge, weaker alternatives can still be used. That is, cases where cooperativity or knowlege are not likely to be given are non-commital to whether stronger alternatives hold. For instance, if for all the speaker knows `Some students came' but she doesn't know whether `All came', then the use of {\em some} succintly conveys her uncertainty about the latter. Thus, given that scalar expressions occur in contexts in which their upper-bounded reading is absent, one could argue for a functional advantage of a lack of semantic upper-bounds: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then these means are preferred over expressing it overtly through a longer expression, e.g. by saying `some but not all' explicitly. Furthermore, not lexicalizing an upper-bound would also be prefered over the hypothetical alternative of lexicalizing two expressions -- one with and one lacking an upper-bound.\footnote{The observation that such monomorphemic expression that lexically rule out stronger alternatives are unattested across languages has received substantial argumentative support (most prominently in \citealt[252-267]{horn:1984} but also e.g. in \citealt{horn:1972,traugott:2004,vdAuwera:2010}). To the best of our knowledge, this claim stands unchallenged.}

In a nutshell, this argument posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so. However, this explanation relies on cues being strongly reliable or morphosyntactic disambiguition being seldom necessary. These assumptions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of safe communication. Two related conditions that might pressure language to English-like semantics are that either morphosyntactic disambiguation is not very costly on its own, or at least less costly than having larger lexica (in this case two for each scalar expression). However, these strike us as technical solutions without a proper empirical basis.

\hl{See notes on Desktop and finish up saying something about probably being a factor, but we see }

\paragraph{Our hyp}
In what follows we investigate the hypothesis that this preference is driven by the advantage in compression that lexical meanings lacking an upper-bound have over those that explicitly codify it.

At present, the contrast in compressability between lexical meanings, however, is not represented as lexica codify the truth-conditions of expressions at a lower level of granularity. \hl{In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.  There is a growing effort to develop such empirically  testable  representational  languages. For  instance, the so-called language of thought has been put to test in various rational probabilistic models that show encouraging results (see e.g. Katz et al. 2008; Piantadosi et al. under review, 2012 and references therein). We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment given that the introduction of a larger framework would also require further assumptions and justifications.}

\subsection{Application}

\paragraph{Procedural description.} The game is initialized with some arbitrary distribution over player types. At the game's onset we compute $Q$ once based on the sets  of sequences $D$ (one for each parent type). Replicator dynamics are computed based on the fitness of each type in the current population as usual. $Q$ is computed anew for each independent run (of $g$ generations) given that it depends on $D$, which is sampled from production probabilities.

\paragraph{Languages.} We consider a population of players with two signaling behaviors, literal and Gricean (level $0$ and $1$ below), each equipped with one of $6$ lexicons. This yields a total of $12$ distinct player types $t \in T$. $|M| = |S| = 2$, i.e., a lexicon is a $(2,2)$-matrix. These are listed in Table \ref{tab:lexica}.

\begin{table}[h]
\centering 
\begin{tabular}{l c l}
$L_1$ = $\begin{pmatrix} 0 & 0 \\ 1 & 1 \end{pmatrix}$ & 
$L_2$ = $\begin{pmatrix} 1 & 1 \\ 0 & 0 \end{pmatrix}$ & 
$L_3$ = $\begin{pmatrix} 1 & 1 \\ 1 & 1 \end{pmatrix}$\\[0.5cm]

@@

As in the CogSci paper, $L_4$ (semantic upper-bound for $m_2$) and $L_5$ (no semantic upper-bound for $m_2$) are the target lexica. Gricean $L_5$ users can convey/infer the bound pragmatically, while literal/Gricean $L_4$ users do so semantically.

@@

\subsection{Analysis}
\hl{Possibly put only fitness and only learning in a single plot with shared y axis}

\begin{figure}
\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{./only-R-or-M}
\end{figure}
%
%\begin{figure}
%\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/07learning}
%\end{figure}

\begin{figure}
\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/03cost-with-l1}
\end{figure}

\begin{figure}
\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/01heatmap}
\end{figure}

\section{Discussion}

\section{Extensions}
\subparagraph{(I) Cost for pragmatic reasoning.} At least in the CogSci setup the effect of adding cost to pragmatic reasoning is unsurprising: High cost for pragmatic signaling lowers the prevalence of pragmatic types. Lexica that semantically encode an upper-bound benefit the most from this. However, the cost needed to be substantial to make the pragmatic English-like lexicon stop being the incumbent type (particularly when learning is communal).

>>>>

We see two main explanations for the lack of upper-bounds in the lexical meaning of weak scalar expressions. The first is that their truth-conditional compatibility with stronger expressions endows them a broader range of application. Scalar expressions occur in contexts in which their upper-bounded reading is absent. This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant. For instance, if for all the speaker knows `Some students came' but she doesn't know whether `All came', then the use of {\em some} succinctly conveys her uncertainty about the latter. This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then these means are preferred over expressing it overtly through a longer expression, e.g., by stating `some but not all' explicitly. Importantly, although dispreferred due to its relative length and complexity, morphosyntactic disambiguation still allows speakers to enforce an upper-bound to override contextual cues that might otherwise mislead the hearer.

In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so. Compare this with a hypothetical language lexicalizes two expressions instead of a single scalar one; one with and one lacking an upper-bound.\footnote{The observation that monomorphemic expression that lexically rule out stronger alternatives are unattested across languages has received substantial argumentative support (most prominently in \citealt[252-267]{horn:1984} but also e.g. in \citealt{horn:1972,traugott:2004,vdAuwera:2010}). To the best of our knowledge, this claim stands unchallenged.} Should the explanation rest on purely functional grounds, we see four conditions that may pressure languages for English-like semantics over this alternative. First, contextual cues are strongly reliable. Second, morphosyntactic disambiguation is seldom necessary. Third, morphosyntactic disambiguation is only marginally dispreferred. Fourth, larger lexica are costly.

However, these conditions are not convincing in their role as central explanatory devices for such a wide-spread phenomenon. The first two put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of . It is likely that human language users are very good at retrieving cues from contexts, but to stipulate that they are so good as to undercut the benefit of safe communication provided by the hypothetical alternative strikes us as too strong of an assumption.  As for conditions (3) and (4), these seem mostly like technical solutions, without a proper empirical basis.

Instead, in what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences is driven by the advantage in compression that lexical meanings lacking an upper-bound have over those that explicitly codify it. Note however that we do not represent this contrast in compression between lexical meanings explicitly in lexica. Instead, the bias towards a lack of upper-bounds in weak scalar alternatives is directly encoded in the learners' prior over types.

 In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.  There is a growing effort to develop such empirically  testable  representational  languages. For  instance, the so-called language of thought has been put to test in various rational probabilistic models that show encouraging results (see e.g. \citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012} and references therein). We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment given that the introduction of a larger framework would also require further assumptions and justifications.\footnote{\hl{T: We could possibly show the length difference of the lexical meanings with and without an upper-bound using a LOT grammar in the appendix. I don't know if it is worth it though.}}

In sum, while we do not want to argue that functionalist pressure may not play a role, we do see a clear benefit in exploring whether matters of learnability would not give us additional leverage.

\subsection{Preliminaries}
The dynamics are initialized with an arbitrary distribution over types. We consider populations with two signaling behaviors, literal or Gricean (level $0$ and $1$ reasoners), each equipped with one of $6$ lexicons. This yields a total of $12$ distinct types $t \in T$. As in our examples above, $|M| = |S| = 2$, i.e., lexica are $(2,2)$-matrices. These are listed in Table \ref{tab:lexica}.

\begin{table}[t]
\centering 
\begin{tabular}{l c l}
$L_1$ = $\begin{pmatrix} 0 & 0 \\ 1 & 1 \end{pmatrix}$ & 
$L_2$ = $\begin{pmatrix} 1 & 1 \\ 0 & 0 \end{pmatrix}$ & 
$L_3$ = $\begin{pmatrix} 1 & 1 \\ 1 & 1 \end{pmatrix}$\\[0.5cm]

@@

Target lexica....

A language that has conventionalized upper and lower bounds to realize a (quasi-)partition of the relevant semantic space will produce speaker behavior that is /almost/ indistinguishable from that of a language with only the respective upper bounds, but with Gricean speakers. Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description. 
\hl{explain why only these and not others}

\begin{table}
\centering
\begin{tabular}{l|l|l}
    \multicolumn{1}{c}{parameter} & \multicolumn{1}{c}{explanation} & \multicolumn{1}{c}{locus}\\ \hline
    $\lambda \geq 1$ & rationality parameter & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\
    $\alpha \in (0,1)$ & semantics-pragmatics tension & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\ 
    $|D|$ & learning data produced per parent type & $P(d|t_j)P(t_i|d)$\\
    $k = |d|$ & number of observations per datum& $P(d|t_j)P(t_i|d)$\\
    $l \geq 1$ & posterior parameter from sampling to MAP & $P(t_i|d) \propto [P(t_i)P(d|t_i)]^l$\\
    $c \in (0,1]$ & learning bias for upper-bound lack &  $P(t_i)$
\end{tabular}
\end{table}

\paragraph{Procedural description.} At the game's onset we compute $Q$ once based on the sets  of sequences $D$ (one for each parent type). Replicator dynamics are computed based on the fitness of each type in the current population as usual. $Q$ is computed anew for each independent run (of $g$ generations) given that it depends on $D$, which is sampled from production probabilities.

@@

\subsection{Analysis}
\begin{figure}
\centering
\includegraphics[scale=.5]{./only-R-or-M}
\end{figure}
\begin{figure}
\centering
\includegraphics[scale=.5]{../presentations/03cost-with-l1}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=.5]{../presentations/01heatmap}
\end{figure}

\subsection{Discussion}

- There is experimental evidence that scalar implicature calculation costs effort and takes additional processing time (Schaeken, Snedeker, ...).

\section{General discussion}

The model combines:
\begin{itemize}
  \item game-theoretical models of functional pressure towards efficient communication
  \item effects of learning biases on (iterated) language learning
  \item assumptions about particular learning biases (Piantadosi et al. 20XX)
  \item probabilistic speaker and listener types of various pragmatic sophistication (Frank \& Goodman, 2012; Franke \& Jäger, 2014)
  \item speaker and listener types with different lexica (Bergen et al. 2012, to appear)
\end{itemize}

Discussion about expressivity as external to learning (cf. Stadler, replicator-papers by Kenny Smith, Kirby et al 2015). Possibly add appendix with direct comparison between IL and RMD.

{\bf Extensions:}

\subparagraph{(I) Cost for pragmatic reasoning.} At least in the CogSci setup the effect of adding cost to pragmatic reasoning is unsurprising: High cost for pragmatic signaling lowers the prevalence of pragmatic types. Lexica that semantically encode an upper-bound benefit the most from this. However, the cost needed to be substantial to make the pragmatic English-like lexicon stop being the incumbent type (particularly when learning is communal).
